A Mechanised Library for Generic (Co)Programming

### 2024-02-27 paper/paper.tex
I did some minor rewriting

\subsection{Functors and Containers}
\label{sec:containers}
A container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
An \emph{extension} of this container is a functor defined as follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms. Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} to the functors defined by a specific
container extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \coq{Valid} that
determines whether a position is valid in a shape.
\begin{coqcode}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{coqcode}
This approach solves the the problem of the unsafe casts, since the code
extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \coq{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \coq{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
Then, we define \emph{container extensions} as
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}

>>>>

\subsection{Functors and Containers}
\label{sec:containers}
A container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
An \emph{extension} of this container is a functor defined as follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms. Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} the functors defined by a specific
container extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \coq{Valid} that
determines whether a position is valid in a shape.
\begin{coqcode}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{coqcode}
This approach solves the problem of the unsafe casts, since the code
extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \coq{Pos}.

Additionaly we would like to reason about equality of
container extensions. The contents of the container extensions need to be
accessed via a function \coq{contents} whose domain depends on the shape of the container.
This means that for a given type of shapes and positions, the equality of two
objects of a container extension will need to deal with heterogeneous equality.
Since we want to remain axiom-free, one way to deal with this is to rely on a
\emph{decidable} validity predicate.

Finally we can define the type of containers we need for performing Coq
extraction correctly. First, containers are defined as pairs of shape and
position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean and lift booleans to propositions similarly to the implementation provided by
SSReflect~\cite{GonthierL09}.
For simplicity, we define the dependent pair of valid positions as a record,
which will be extracted to a singleton type with the validity proof erased by
Coq's code extraction mechanism.
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
Then, we define \emph{container extensions} as follows:
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}
